---
title: "Take-home Labs"
output:
  html_document:
    toc: false
---

```{r setup, include=FALSE, echo=FALSE, results='hide'}
# List of all required packages across labs
required_packages <- c(
  "tidyverse",     # Used throughout for data manipulation and plotting
  "lubridate",     # Lab 8 - date/time handling
  "viridis",       # Lab 7 - color scales
  "purrr",         # Lab 9 - iteration and mapping
  "ggplot2",       # Used throughout for plotting (part of tidyverse but sometimes loaded separately)
  "dplyr",         # Used throughout for data manipulation (part of tidyverse but sometimes loaded separately)
  "readr"          # Used throughout for data import (part of tidyverse but sometimes loaded separately)
)

# Check which packages need to be installed
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]

# Install missing packages
if(length(new_packages)) {
  install.packages(new_packages)
}

# Load all required packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

Labs are designed to reinforce the code/lessons covered that week and provide you a chance to practice working with R. Labs are to be completed in your local project (on your computer) and uploaded to Canvas by the beginning of class (1pm) the following Thursday. That said, these due dates are largely suggestive as a way to help you prioritize and stay caught up as a group -- if you need, or want, more time, take it. At the end of the quarter, we will simply look over the tasks you have completed in concert with the reflection you submit.


<br>

## Labs {.tabset .tabset-fade .tabset-pills}

### <small>Lab Week 1</small> {#lab-week-1}

This lab will help you get oriented with R, RStudio, and your first programming concepts. The goal is to practice the fundamentals covered in the the introduction to R and RStudio lecture.

**Setup Instructions:**

1. Open RStudio

2. Create a new R script by going to `File -> New File -> R Script` or pressing `Ctrl/Cmd+Shift+N`

3. Save your script as `lastname_lab_week_1.R` in an appropriate location (with your own last name)

4. As you work through the problems below, write your code in the script and run it using `Ctrl/Cmd+Enter` (try to practice not highlighting and clicking "run")

**Part 1: Basic Arithmetic and Order of Operations**

1. Use parentheses to edit `15 + 7 * 3` force R to return `66` instead of `36`.

2. Calculate the square root of 144.

3. Calculate 2 raised to the power of 8.

**Part 2: Objects and Assignment**

4. Create an object called `current_year` and assign a value of `2025` to it using the `<-` operator.

5. Calculate how many years it has been since `1977` using the `current_year` object.

**Part 3: Mathematical Functions**

6. Calculate the natural logarithm of 10 using the `log()` function.

7. Calculate e raised to the power of 2 using the `exp()` function. 


**Part 4: Logical Comparisons**

8. Test whether 10 is equal to 5 + 5 using the `==` operator.

9. Test whether 7 is greater than 10.

10. Test whether 15 is greater than or equal to 15.

11. Create an object called `x` with the value 25, then test whether `x` is not equal to 30.

**Part 5: Exploring Functions and Getting Help**

12. Use the help function to learn about the `round()` function by typing `?round` in the console.

13. Round the number 3.14159 to 2 decimal places using the `round()` function.

14. Calculate the absolute value of -15 using the `abs()` function.

**Part 6: Working with Scripts** 

15. Add comments to your script using `#` to explain what each section does.

16. Save your script and make sure you can run the entire script from top to bottom without errors.


### <small>Lab Week 2</small> {#lab-week-2}

In this lab we are going to practice subsetting and manipulating vectors.  

First, open a new script in your workgin directory and save it to your `scripts` folder. Call this new script `lastname_week_2_lab` (with your last name). 

Copy and paste the chunk of code below into your new `lastname_week_2_lab` script and run it. This chunk of code will create the vector you will use in your lab today. Check in your environment to see what it looks like. What do you think each line of code is doing? 

```{r}
set.seed(15)
hw2 <- runif(50, 4, 50)
hw2 <- replace(hw2, c(4,12,22,27), NA)
hw2
```


1. Take your `hw2` vector and removed all the NAs then select all the numbers between 14 and 38 inclusive, call this vector `prob1`. 

2. Multiply each number in the `prob1` vector by 3 to create a new vector called `times3`. Then add 10 to each number in your `times3` vector to create a new vector called `plus10`. 

3. Select every other number in your `plus10` vector by selecting the first number, not the second, the third, not the fourth, etc. If you've worked through these three problems in order, you should now have a vector that is 12 numbers long that looks **exactly** like this one:

```{r, echo = F}
prob1 <- hw2[!is.na(hw2)] #removing the NAs

prob1 <- prob1[prob1 >14 & prob1 < 38] #only selecting numbers between 14 and 38

times3 <- prob1 * 3 #multiplying by 3

plus10 <- times3 + 10 #adding 10 to the whole vector 

final <- plus10[c(TRUE, FALSE)] #selecting every other number
```

```{r}
final
```

Finally, save your script and upload your file to Canvas. 




### <small>Lab Week 3</small> {#lab-week-3}

Lab this week will be playing with the `surveys` data we worked on in class. Create a new script within your  `scripts` folder called `lastname_week_3_lab.R` (with your own last name). 

Load the `surveys` data frame with the read.csv() function. Create a new data frame called `surveys_base` with only the species_id, the weight, and the plot_type columns. Have this data frame only be the first 5,000 rows. Convert both species_id and plot_type to factors. Remove all rows where there is an NA in the weight column. Explore these variables and try to explain why a factor is different from a character. Why might we want to use factors? Can you think of any examples? 

CHALLENGE:
Create a second data frame called `challenge_base` that only consists of individuals from your `surveys_base` data frame with weights greater than 150g.  



<br> 

### <small>Lab Week 4</small> {#lab-week-4}

This week the lab will review data manipulation in the tidyverse.

1. Create a tibble named `surveys` from the portal_data_joined.csv file, found at https://ucd-rdavis.github.io/R-DAVIS/data/portal_data_joined.csv.

2. Subset `surveys` using Tidyverse methods to keep rows with weight between 30 and 60, and print out the first 6 rows.

3. Create a new tibble showing the maximum weight for each species + sex combination and name it `biggest_critters`. Sort the tibble to take a look at the biggest and smallest species + sex combinations. HINT: it's easier to calculate max if there are no NAs in the dataframe...

4. Try to figure out where the NA weights are concentrated in the data- is there a particular species, taxa, plot, or whatever, where there are lots of NA values? There isn’t necessarily a right or wrong answer here, but manipulate surveys a few different ways to explore this. Maybe use `tally` and `arrange` here.

5. Take `surveys`, remove the rows where weight is NA and add a column that contains the average weight of each species+sex combination to the **full** `surveys` dataframe. Then get rid of all the columns except for species, sex, weight, and your new average weight column. Save this tibble as `surveys_avg_weight`.

6. Take `surveys_avg_weight` and add a new column called `above_average` that contains logical values stating whether or not a row’s weight is above average for its species+sex combination (recall the new column we made for this tibble).



### <small>Lab Week 5</small> {#lab-week-5}

This week's questions will have us practicing pivots and conditional statements.

1. Create a tibble named `surveys` from the portal_data_joined.csv file, found at https://ucd-rdavis.github.io/R-DAVIS/data/portal_data_joined.csv. Then manipulate `surveys` to create a new dataframe called `surveys_wide` with a column for genus and a column named after every plot type, with each of these columns containing the mean hindfoot length of animals in that plot type and genus. So every row has a genus and then a mean hindfoot length value for every plot type. The dataframe should be sorted by values in the Control plot type column. This question will involve quite a few of the functions you've used so far, and it may be useful to sketch out the steps to get to the final result.

2. Using the original `surveys` dataframe, use the two different functions we laid out for conditional statements, ifelse() and case_when(), to calculate a new weight category variable called `weight_cat`. For this variable, define the rodent weight into three categories, where "small" is less than or equal to the 1st quartile of weight distribution, "medium" is between (but not inclusive) the 1st and 3rd quartile, and "large" is any weight greater than or equal to the 3rd quartile. (Hint: the summary() function on a column summarizes the distribution). For ifelse() and case_when(), compare what happens to the weight values of NA, depending on how you specify your arguments.

BONUS: How might you soft code the values (i.e. not type them in manually) of the 1st and 3rd quartile into your conditional statements in question 2?



### <small>Lab Week 7</small> {#lab-week-7}

For our week seven lab, we are going to be practicing the skills we learned with ggplot during class. You will be happy to know that we are going to be using a brand new data set called `gapminder`. This data set is looking at statistics for a few different counties including population, GDP per capita, and life expectancy. Download the data using the code below.

```{r, warnings = F}
library(tidyverse)

gapminder <- read_csv("https://ucd-rdavis.github.io/R-DAVIS/data/gapminder.csv") #ONLY change the "data" part of this path if necessary

```

**Part A: Basic ggplot Skills**

1. First calculates mean life expectancy on each continent. Then create a plot that shows how life expectancy has changed over time in each continent. Try to do this all in one step using pipes! (aka, try not to create intermediate dataframes)

2. Look at the following code and answer the following questions. What do you think the `scale_x_log10()` line of code is achieving? What about the `geom_smooth()` line of code? 

*Challenge!* Modify the above code to size the points in proportion to the population of the country. 
**Hint:** Are you translating data to a visual feature of the plot?

**Hint:** There's no cost to tinkering! Try some code out and see what happens with or without particular elements.

```{r}

ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point(aes(color = continent), size = .25) + 
    scale_x_log10() +
    geom_smooth(method = 'lm', color = 'black', linetype = 'dashed') +
    theme_bw()

```

3. Create a boxplot that shows the life expectency for Brazil, China, El Salvador, Niger, and the United States, with the data points in the backgroud using geom_jitter. Label the X and Y axis with "Country" and "Life Expectancy" and title the plot "Life Expectancy of Five Countries".

**Part B: Advanced ggplot Skills - Graph Recreation**

For the second part of this lab, we're going to be working on 2 critical `ggplot` skills: recreating a graph from a dataset and **googling stuff**.

Our goal will be to make this final graph using the `gapminder` dataset:

```{r, eval=T, echo=F, warning=F, message=F}
library(tidyverse)

gapminder <- read_csv("https://ucd-rdavis.github.io/R-DAVIS/data/gapminder.csv")

pg <- gapminder %>% 
  select(country, year, pop, continent) %>% 
  filter(year > 2000) %>% 
  pivot_wider(names_from = year, values_from = pop) %>% 
  mutate(pop_change_0207 = `2007` - `2002`)
```

```{r, eval=T, echo=F, warning=F, message=F, fig.width = 8, fig.height = 5}
pg %>% 
  filter(continent != "Oceania") %>% 
  ggplot(aes(x = reorder(country, pop_change_0207), y = pop_change_0207)) +
  geom_col(aes(fill = continent)) +
  facet_wrap(~continent, scales = "free") +
  theme_bw() +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none") +
  xlab("Country") +
  ylab("Change in Population Between 2002 and 2007")

```

The x axis labels are all scrunched up because we can't make the image bigger on the webpage, but if you make it and then zoom it bigger in RStudio it looks much better.

We'll touch on some intermediate steps here, since it might take quite a few steps to get from start to finish. Here are some things to note:

1. To get the population difference between 2002 and 2007 for each country, it would probably be easiest to have a country in each row and a column for 2002 population and a column for 2007 population.

2. Notice the order of countries within each facet. You'll have to look up how to order them in this way.

3. Also look at how the axes are different for each facet. Try looking through `?facet_wrap` to see if you can figure this one out.

4. The color scale is different from the default- feel free to try out other color scales, just don't use the defaults!

5. The theme here is different from the default in a few ways, again, feel free to play around with other non-default themes.

6. The axis labels are rotated! Here's a hint: `angle = 45, hjust = 1`. It's up to you (and Google) to figure out where this code goes!

7. Is there a legend on this plot?

This lesson should illustrate a key reality of making plots in R, one that applies as much to experts as beginners: 10% of your effort gets the plot 90% right, and 90% of the effort is getting the plot perfect. `ggplot` is incredibly powerful for exploratory analysis, as you can get a good plot with only a few lines of code. It's also extremely flexible, allowing you to tweak nearly everything about a plot to get a highly polished final product, but these little tweaks can take a lot of time to figure out!

So if you spend most of your time on this lesson googling stuff, you're not alone!





### <small>Lab Week 8</small> {#lab-week-8}

Let's look at some real data from Mauna Loa to try to format and plot. These meteorological data from Mauna Loa were collected every minute for the year 2001. *This dataset has 459,769 observations for 9 different metrics of wind, humidity, barometric pressure, air temperature, and precipitation.*  You can read the CSV directly from the R-DAVIS Github:
`mloa <- read_csv("https://ucd-rdavis.github.io/R-DAVIS/data/mauna_loa_met_2001_minute.csv")`

Use the [README](https://ucd-rdavis.github.io/R-DAVIS/data/mauna_loa_README.txt) file associated with the Mauna Loa dataset to determine in what time zone the data are reported, and how missing values are reported in each column. With the `mloa` data.frame, remove observations with missing values in rel_humid, temp_C_2m, and windSpeed_m_s. Generate a column called "datetime" using the year, month, day, hour24, and min columns. Next, create a column called "datetimeLocal" that converts the datetime column to Pacific/Honolulu time (*HINT*: look at the lubridate function called `with_tz()`). Then, use dplyr to calculate the mean hourly temperature each month using the temp_C_2m column and the datetimeLocal columns. (*HINT*: Look at the lubridate functions called `month()` and `hour()`). Finally, make a ggplot scatterplot of the mean monthly temperature, with points colored by local hour.


### <small>Lab Week 9</small> {#lab-week-9}
In this assignment, you'll use the iteration skills we built in the course to apply functions to an entire dataset.

Let's load the surveys dataset:

```{r wk9a}
surveys <- read.csv("https://ucd-rdavis.github.io/R-DAVIS/data/portal_data_joined.csv")
```

1) Using a for loop, print to the console the longest species name of each taxon. Hint: the function nchar() gets you the number of characters in a string.

Next let's load the Mauna Loa dataset from last week.

```{r week9b}
mloa <- read_csv("https://ucd-rdavis.github.io/R-DAVIS/data/mauna_loa_met_2001_minute.csv")
```

2) Use the map function from purrr to print the max of each of the following columns: "windDir","windSpeed_m_s","baro_hPa","temp_C_2m","temp_C_10m","temp_C_towertop","rel_humid",and "precip_intens_mm_hr". 

3) Make a function called C_to_F that converts Celsius to Fahrenheit. Hint: first you need to multiply the Celsius temperature by 1.8, then add 32. Make three new columns called "temp_F_2m", "temp_F_10m", and "temp_F_towertop" by applying this function to columns "temp_C_2m", "temp_C_10m", and "temp_C_towertop". Bonus: can you do this by using map_df? Don't forget to name your new columns "temp_F..." and not "temp_C..."!

Challenge: Use lapply to create a new column of the `surveys` dataframe that includes the genus and species name together as one string.



