---
title: "Wk08: Dates & Times with Lubridate and Creating Functions"
subtitle: Working with datetimes and writing your own functions
minutes: 30
---

<div class = "blue">
### Learning Objectives
* Learn the basic date/datetime types in R
* Gain familiarity with converting dates and timezones
* Learn how to use the `lubridate` package 
* Tips and tricks about management of datetime data
* Define a function that takes arguments.
* Set default values for function arguments.
* Explain why we should divide programs into small, single-purpose functions. 

</div>
<br>

## Videos

- Dates with Lubridate (coming soon)
- Function-writing (coming soon)
- [Video Discussion](https://canvas.ucdavis.edu/courses/1019595/assignments/1510685)

## In Class Coding

[Submit Week 8 In Class Coding](https://canvas.ucdavis.edu/courses/1019595/assignments/1509413)

# Add a package check/install code block after the YAML header.

```{r setup, include=FALSE, echo=FALSE, results='hide'}
# List of all required packages across assignments
required_packages <- c(
  "lubridate",
  "tidyverse",     # Used throughout for data manipulation and plotting
  "viridis",       # Assignment 7 - color scales
  "purrr",         # Assignment 9 - iteration and mapping
  "ggplot2",       # Used throughout for plotting (part of tidyverse but sometimes loaded separately)
  "dplyr",         # Used throughout for data manipulation (part of tidyverse but sometimes loaded separately)
  "readr"          # Used throughout for data import (part of tidyverse but sometimes loaded separately)
)
```

<br>

# Part 1: Date-Times in R

We have learned about different data type classes in previous lessons. Some common data classes we have examined before include character, factor, and numeric. But R also recognizes a data class called "Dates". Having your date data in the "Dates" data class is very useful, as you can then do things like calculate time between two events, transform the dates into different formats, and plot temporal data easily. 
In this lesson, we are going to introduce how base R deals with dates (`POSIXct` or `POSITlt`), but we are going to spend the majority of our lesson on the package `lubridate`. `lubridate` is a great package that makes it much easier to work with dates and times in R. 

## Date-Time Classes in Base R

Importantly, there are **3** basic time classes in R:

 - **`Dates`** (just dates, i.e., 2012-02-10)
 - **`POSIXct`** ("**ct**" == calendar time, best class for dates with times)
 - **`POSIXlt`** ("**lt**" == local time, enables easy extraction of specific components of a time, however, remember that POXIXlt objects are lists)

Unfortunately converting dates & times in R into formats that are computer readable can be frustrating, mainly because there is very little consistency. In particular, if you are importing things from Excel, keep in mind dates can get especially weird^1^, depending on the operating system you are working on, the format of your data, etc.

^*1*^ <sub>*For example Excel stores dates as a number representing days since 1900-Jan-0, plus a fractional portion of a 24 hour day (**serial-time**), but in OSX (Mac), it is 1904-Jan-0.*</sub>

## Dates

The `Date` class in R can easily be converted or operated on numerically, depending on the interest. Let's make a string of dates to use for our example:

```{r}
sample_dates_1 <- c("2018-02-01", "2018-03-21", "2018-10-05", "2019-01-01", "2019-02-18")
#notice we have dates across two years here
```

What is the class that R classifies this data as? 

R classifies our `sample_dates_1` data as character data. Let's transform it into Dates. Notice that our `sample_dates_1` is in a nice format: YYYY-MM-DD. This is the format necessary for the function `as.Date`.

```{r}

sample_dates_1 <- as.Date(sample_dates_1)

```

What happens with different orders...say MM-DD-YYYY? 

```{r sampledates, echo=T, eval=T}

# Some sample dates: 

sample_dates_2 <- c("02-01-2018", "03-21-2018", "10-05-2018", "01-01-2019", "02-18-2019")

sample_dates_3 <-as.Date(sample_dates_2) # well that doesn't work

```

The reason this doesn't work is because the computer expects one thing, but is getting something else. Remember, **write code you can read and your computer can understand**. So we need to give some more information here so R will interpret our data correctly.

```{r sampledates2, echo=T, eval=T}

# Some sample dates:
sample_dates_2 <- c("02-01-2018", "03-21-2018", "10-05-2018", "01-01-2019", "02-18-2019")

sample_dates_3<- as.Date(sample_dates_2, format = "%m-%d-%Y" ) # date code preceded by "%"

```


To see a list of the date-time format codes in R, check out this [page and table](https://www.stat.berkeley.edu/~s133/dates.html), or you can use: `?(strptime)`

The nice thing is this method works well with pretty much any format, you just need to provide the associated codes and structure:

 - `as.Date("2016/01/01", format="%Y/%m/%d")`=`r as.Date("2016/01/01", format="%Y/%m/%d")`
 
 - `as.Date("05A21A2011", format="%mA%dA%Y")`=`r as.Date("05A21A2011", format="%mA%dA%Y")`
 
<div class = "blue">
### Challenge

Format this date with the `as.Date` function: `Jul 04, 2019`

<details>
<summary>ANSWER</summary>
```{r challenge1, purl=FALSE}

as.Date("Jul 04, 2019", format = "%b%d,%Y")
```
</details>
</div>

<br>

## Working with Times in Base R

When working with times, the best class to use in base R is `POSIXct`. 


```{r}

tm1 <- as.POSIXct("2016-07-24 23:55:26")
tm1
tm2 <- as.POSIXct("25072016 08:32:07", format = "%d%m%Y %H:%M:%S")
tm2

#Notice how POSIXct automatically uses the timezone your computer is set to. What if we collected this data in a different timezone?

# specify the time zone:
tm3 <- as.POSIXct("2010-12-01 11:42:03", tz = "GMT")
tm3

```


## The `lubridate` Package

The `lubridate` package will handle 90% of the date & datetime issues you need to deal with. It is fast, much easier to work with, and we recommend using it wherever possible. Do keep in mind sometimes you need to fall back on the base R functions (i.e., `as.Date()`), which is why having a basic understanding of theses functions and their use is important. 

To use `lubridate` we will first need to install and load the package.

```{r, message = F}
#install.packages("lubridate")

library(lubridate)
```

`lubridate` has lots of handy functions for converting between date and time formats, and even timezones. 

Let's take a look at our `sample_dates_1` data again.

```{r}
sample_dates_1 <- c("2018-02-01", "2018-03-21", "2018-10-05", "2019-01-01", "2019-02-18")
```
Once again, R reads this in a character data. 

Lubridate uses functions that looks like `ymd` or `mdy` to transform data into the class "Date". Our `sample_dates_1` data is formatted like Year, Month, Day, so we would use the `lubridate` function `ymd` (y = year, m = month, d = day). 

```{r}
sample_dates_lub <- ymd(sample_dates_1)
```

What about that messier `sample_dates_2` data? Remember R wants dates to be in the format YYYY-MM-DD. 

```{r}
sample_dates_2 <- c("2-01-2018", "3-21-2018", "10-05-18", "01-01-2019", "02-18-2019")
#notice that some numbers for years and months are missing

sample_dates_lub2 <- mdy(sample_dates_2) #lubridate can handle it! 
```

All sorts of date formats can easily be transformed using `lubridate`:

 - `lubridate::ymd("2016/01/01")`=`r lubridate::ymd("2016/01/01")`
 - `lubridate::ymd("2011-03-19")`=`r lubridate::ymd("2011-03-19")`
 - `lubridate::mdy("Feb 19, 2011")`=`r lubridate::mdy("Feb 19, 2011")`
 - `lubridate::dmy("22051997")`=`r lubridate::dmy("22051997")`


## Using `lubridate` for Time and Timezones

`lubridate` has very similar functions to handle data with Times and Timezones. To the `ymd` function, add `_hms` or `_hm` (h= hours, m= minute, s= seconds) and a `tz` argument. `lubridate` will default to the POSIXct format.

 - Example 1: `lubridate::ymd_hm("2016-01-01 12:00", tz="America/Los_Angeles")` = `r lubridate::ymd_hm("2016-01-01 12:00", tz="America/Los_Angeles")`
 - Example 2 (24 hr time): `lubridate::ymd_hm("2016/04/05 14:47", tz="America/Los_Angeles")` = `r lubridate::ymd_hm("2016/04/05 14:47", tz="America/Los_Angeles")`
 - Example 3 (12 hr time but converts to 24): `lubridate::ymd_hms("2016/04/05 4:47:21 PM", tz="America/Los_Angeles")` = `r lubridate::ymd_hms("2016/04/05 4:47:21 PM", tz="America/Los_Angeles")`


### Lubridate Tips

For lubridate to work, you need the column datatype to be **character** or **factor**. The `readr` package (from the `tidyverse`) is smart and will try to guess for you. Problem is, it might convert your data for you without the settings (in this case the proper timezone). So here are few ways to work around this. 

```{r, message = F, warning = F}
library(lubridate)
library(dplyr)
library(readr)

# read in some data and skip header lines
nfy1 <- read_csv("data/2015_NFY_solinst.csv", skip = 12)
head(nfy1) #R tried to guess for you that the first column was a date and the second a time

# import raw dataset & specify column types
nfy2 <- read_csv("data/2015_NFY_solinst.csv", col_types = "ccidd", skip=12)

glimpse(nfy1) # notice the data types in the Date.Time and datetime cols
glimpse(nfy2)
```

Next we want to create a single datetime column. How do we get our Date and Time columns into one column so we can format it as a datetime? The answer is the `paste` function.

 - `paste()` allows pasting text or vectors (& columns) by a given separator that you specify with the `sep = ` argument
 - `paste0()` is the same thing, but defaults to using no separator (i.e. no space).

```{r, message = F, warning = F}
# make a datetime col:
nfy2$datetime <- paste(nfy2$Date, " ", nfy2$Time, sep = "")

glimpse(nfy2) #notice the  datetime column is classifed as character

# convert Date Time to POSIXct in local timezone using lubridate
nfy2$datetime_test <- as_datetime(nfy2$datetime, 
                                    tz="America/Los_Angeles")
# OR convert using the ymd_functions
nfy2$datetime_test2 <- ymd_hms(nfy2$datetime, tz="America/Los_Angeles")

# OR wrap in as.character()
nfy1$datetime <- ymd_hms(as.character(paste0(nfy1$Date, " ", nfy1$Time)), tz="America/Los_Angeles")
tz(nfy1$datetime)

```

Last, `lubridate` lets you extract components of date, time and datetime data types with intuitive functions. 

```{r}
# Functions called day(), month(), year(), hour(), minute(), second(), etc... will extract those elements of a datetime column.
months <- month(nfy2$datetime)
# Use the table function to get a quick summary of categorical variables
table(months)

# Add label and abbr agruments to convert numeric representations to have names
months <- month(nfy2$datetime, label = TRUE, abbr=TRUE)
table(months)

```

# Part 2: Creating Functions

Any operation you will perform more than once can be put into a function. That way, rather than retyping all the commands (and potentially making errors), you can simply call the function, passing it a new dataset or parameters. This may seem cumbersome at first, but writing functions to automate repetitive tasks is incredibly powerful. E.g. each time you call `ggplot` you are calling a function that someone wrote. Imagine if each time you wanted to make a plot you had to copy and paste or write that code from scratch!


## Defining a function

Recall the components of a function. E.g. the `log` function (see `?log`) takes "arguments" `x` and `base` and "returns" the base-`base` logarithm of `x`. Functions take arguments as input and yield return-values as output. You can define functions to do any number of operations on any number of arguments, but always output a single return value (however there are complex objects into which you can put multiple objects, should you need to).

Let's start by defining a simple function to add two numbers. This is the basic structure, which you can read as "assign to the variable `my_sum` a function that takes arguments `a` and `b` and returns `the_sum`." The body of the function is delimited by the curly-braces. The statements in the body are indented. This makes the code easier to read but does not affect how the code operates. 

```{r}
my_sum <- function(a, b) {
  the_sum <- a + b
  return(the_sum)
}
```

Notice that no numbers were summed when we ran that code, but now the Environment has an object called `my_sum` that has type function. You can call `my_sum` just like you would any other function. When you do, the code between the curly-braces of the `my_sum` definition is run with whatever values you pass to `a` and `b` substituted in their place.

```{r}
my_sum(a = 2, b = 2)
my_sum(3, 4)
```

Just like `log` provides a default value of `base` (`exp(1)`) so that you don't have to type it every time, you can provide default values to any arguments of your function. Then if the user doesn't specify them, the defaults will be used.

```{r}
my_sum2 <- function(a = 1, b = 2) {
  the_sum <- a + b
  return(the_sum)
}
my_sum2()
my_sum2(b = 7)
```

<div class = "blue">

### Tip

One feature unique to R is that the return statement is not required.
R automatically returns the output of the last line of the body
of the function unless a `return` statement is specified elsewhere.
Since other languages require a `return` statement and because it can make
reading a function easier, we will explicitly define the return statement.
</div>
<br>

## Temperature conversion

Let's define a function F_to_K that converts temperatures from Fahrenheit to Kelvin:

```{r}
F_to_K <- function(temp) {
  K <- ((temp - 32) * (5 / 9)) + 273.15
  return(K)
}
```

Calling our own function is no different from calling any other function:

```{r}
# freezing point of water
F_to_K(32)
```

```{r}
# boiling point of water
F_to_K(212)
```

<div class = "blue">

### Challenge

- Write a function called `K_to_C` that takes a temperature in K
and returns that temperature in C
   - Hint: To convert from K to C you subtract 273.15
- Create a new R script, copy `F_to_K` and `K_to_C` in it, and save it as 
functions.R in the `code` directory of your project.

<details>
<summary>ANSWER</summary>
```{r}
K_to_C <- function(tempK) {
    tempC <- tempK - 273.15 
    return(tempC)
}
```
</details>
</div>
<br>


## `source()`ing functions

You can load all the functions in your `code/functions.R` script without even opening the file, via the `source` function. This allows you to keep your functions separate from the analyses which use them. 

```{r, eval = FALSE}
source('code/functions.R')
```


### Using dataframes in functions

Let's write a function to calculate the average GDP in a given country, in a given span of years, based on the `gapminder` data. If you were to do this for just one specification, without writing a function, it might look something like this:

```{r}
library(gapminder)
library(tidyverse)

gapminder %>% 
  filter(country == "Canada", year %in% c(1950:1970)) %>% 
  summarize(mean(gdpPercap, na.rm = T))
```

But, what if you wanted to do this for many different specifications? You might find yourself wanting to copy and paste these couple lines of code over and over. Instead, you can write it into a function, **soft coding** the parts that you want as your function's arguments. In this example, we want to be able to change the country and the year range that we are interested in. 

```{r}
# Note: try to name arguments something that do not exist as a column name, to avoid confusing yourself and R
avgGDP <- function(cntry, yr.range){
  df <- gapminder %>% 
  filter(country == cntry, year %in% yr.range)
  mean(df$gdpPercap, na.rm = T)
}

avgGDP("Iran", 1980:1985)
avgGDP("Zimbabwe", 1950:2000)

```


## Pass by value

Functions in R almost always make copies of the data to operate on
inside of a function body. When we modify a data frame inside the function
we are modifying the copy of the gapminder dataset,
not the original variable we gave as the first argument.
This is called "pass-by-value" and it makes writing code much safer:
you can always be sure that whatever changes you make within the
body of the function, stay inside the body of the function.


<div class = "blue"> 

### Challenge{.challenge}

This challenge will deal with countries' population growth. To access the data, load (and install, if needed) the `gapminder` library and access its life expectancy dataset using:

```
library(gapminder)
d <- gapminder::gapminder
```

Write a new function that takes two arguments, the gapminder data.frame (`d`) and the name of a country (e.g. `"Afghanistan"`), and plots a time series of the country's population. The return value from the function should be a ggplot object. Note: It is often easier to modify existing code than to start from scratch. To start out with one plot for a particular country, figured out what you need to change for each iteration (these will be your arguments), and then wrap it in a function.  



<details>
<summary>ANSWER</summary>
```{r}
plotPopGrowth <- function(countrytoplot, dat = gapminder) {
    df <- filter(dat, country == countrytoplot) 
    plot <- ggplot(df, aes(year, pop)) + 
        geom_line()
    return(plot)
}
plotPopGrowth('Canada')
```
</details>
</div>
<br>

This lesson was contributed by [Ryan Peek](https://ryanpeek.github.io/) and [Martha Zillig](https://github.com/MarthaWohlfeil).

This lesson is adapted from the Software Carpentry: R for Reproducible Scientific Analysis [Creating Functions materials](http://data-lessons.github.io/gapminder-R/07-functions.html).