---
title: "Lab Week 10"
format:
  html:
    embed-resources: true
---

This is the Lab Week 10 document for R-DAVIS. Complete the tasks below in this file, then click **Render** and upload the resulting HTML file to Canvas.

In this assignment, you'll use the iteration skills we built in the course to apply functions to an entire dataset.

Let's load the surveys dataset:

```{r wk9a}
library(tidyverse)
surveys <- read_csv("https://ucd-rdavis.github.io/R-DAVIS/data/portal_data_joined.csv")
```

1) Using a for loop, print to the console the longest species name of each taxon. Hint: the function nchar() gets you the number of characters in a string. The function unique() gets you the unique values in a vector.

<!-- INSTRUCTOR TODO (2026 reorg): the Mauna Loa dates lab was archived; "last week" no longer refers to it. -->
Next let's load the Mauna Loa dataset.

```{r week9b}
mloa <- read_csv("https://ucd-rdavis.github.io/R-DAVIS/data/mauna_loa_met_2001_minute.csv")
```

2) Use the `apply` function to print the max of each of the following columns: "windDir","windSpeed_m_s","baro_hPa","temp_C_2m","temp_C_10m","temp_C_towertop","rel_humid",and "precip_intens_mm_hr". 

3) Make a function called C_to_F that converts Celsius to Fahrenheit. Hint: first you need to multiply the Celsius temperature by 1.8, then add 32. Make three new columns called "temp_F_2m", "temp_F_10m", and "temp_F_towertop" by applying this function to columns "temp_C_2m", "temp_C_10m", and "temp_C_towertop". 

4) Use the `apply` function to create a new column of the `surveys` dataframe that includes the genus and species name together as one string. Hint: the function we've been using to combine strings together has a "collapse" argument. Do some poking around in the help file to get started.
